---
title: "Saratoga Houses"
author: "Dimitris Athanasakoglou"
date: "July 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries}
load.libraries <- c('corrplot', 'ggplot2', 'dplyr', 'psych')
install.lib <- load.libraries[!load.libraries %in% installed.packages()]
for(libs in install.lib) install.packages(libs, dependencies = TRUE)
sapply(load.libraries, require, character = TRUE)
```


```{r, echo=FALSE}
saratoga<-read.csv("data/SaratogaHouses.csv", header=TRUE)
saratoga<-saratoga[,-1]
```


##Structure of the Data
The housing dataset has 1728 rows and 16 variables with price being the DV. There are no missing values in the data.
```{r structure}
dim(saratoga)
str(saratoga)
sum(is.na(saratoga))
```


##Data Issues
The variable bathrooms has numeric value, with half bathrooms indicating bathrooms without bathtubs.
```{r missing data}
is.numeric(saratoga$bathrooms)
```

##Descriptive Statistics

```{r descriptive stats}
summary(saratoga)
```

##Internal House Characteristics
Correlations among internal house charcteristics and price. The DV (price) has many strong positive correlations with many IVs.
```{r internal characteristics}
internal_chars <- c('age', 'lotSize', 'landValue', 'livingArea', 'bedrooms', 'fireplaces', 'bathrooms', 'rooms', 'pctCollege', 'price')
corrplot(cor(saratoga[, internal_chars]))
```


Price has the strongest correlation with livingArea. We further analyse the relation between the variables using a scatterplot. As we can see there is a linear positive correlation with many outliers. 
```{r price livingArea}
ggplot(aes(x=livingArea, y=price), data=saratoga) + geom_point(color='blue')
```


##Visualise Without the Outliers (exclude houses >3000 sq feet).
```{r exclude outliers}
saratoga %>% 
        filter(livingArea < 3000) %>%
        ggplot(aes(x=livingArea, y=price)) + 
        geom_point(color='blue', alpha=0.5) +
        labs(x='Area', y='Price', title='Price by living area in sq feet')
```


##Analyse Rooms
Strong positive correlations between rooms, bathrooms, bedrooms (expected).
```{r rooms}
rooms <- c('bedrooms', 'bathrooms', 'rooms')
corrplot(cor(saratoga[, rooms]))
```


##Room Distribution
The majority of the houses have 7 rooms or less.
```{r room distribution}
table(saratoga$rooms)
ggplot(aes(x=rooms), data=saratoga) + 
        geom_histogram(fill='blue', bins=15, binwidth = 1) + 
        ggtitle('Distribution of room count')
```


##Age Distribution
There are 83 observations with age zero and many outliers with only one house attached. The distribution has positive skewness and the extreme majority of houses are below 50 years old.
```{r age distribution}
table(saratoga$age)
saratoga %>% 
        filter(age <= 130) %>% 
        ggplot(aes(x=age)) + 
        geom_histogram(fill='blue') + 
        ggtitle('Distribution of house age')
```


##Checking the Relation Between Price and Age
The relationship appears somewhat steady over time, especially for houses below 50 years. There is strong volatility in the later years. This is not a real effect but simply due to the sparseness of observations for houses above 50 years old, as the age distribution revealed.
```{r price and age}
saratoga %>% 
        filter(age <= 130) %>%
        group_by(age) %>% 
        summarize(mean_price=mean(price)) %>%
        ggplot(aes(x=age, y=mean_price)) +
        geom_line(stat='identity', color='blue') + 
        geom_smooth(color='darkgrey') +
        ggtitle('Mean price by age')
```


##Analyse the prices
To check the normality to the distributions of the variables we check the histograms.
```{r DV distr}
#price
saratoga %>% 
        ggplot(aes(x=price)) + 
        geom_histogram(fill='blue') + 
        ggtitle('Distribution of DV price')

#price log10
saratoga %>% 
        ggplot(aes(x=log10(price))) + 
        geom_histogram(fill='blue') + 
        ggtitle('Distribution of DV price')
```


##Multi Histograms
```{r Histograms}
#non-normalized
saratoga %>%
        select(price, lotSize, age, livingArea, pctCollege,bedrooms,
               fireplaces, bathrooms, rooms) %>%
        multi.hist(density=TRUE, dcol="red", main = "Histogram")

#normalized histograms     
saratoga %>%
        select(price, lotSize, age, livingArea, pctCollege,bedrooms,
               fireplaces, bathrooms, rooms) %>%
        sapply(log) %>%
        multi.hist(density=TRUE, dcol="red", main = "Histogram")
```


##Analysis of Variance (ANOVA)
```{r anova}
anova1 <- aov(saratoga$price ~ saratoga$sewer)
anova1
TukeyHSD(anova1)

anova2 <- aov(saratoga$price ~ saratoga$fuel)
anova2
TukeyHSD(anova2)

anova3 <- aov(saratoga$price ~ saratoga$heating)
anova3
TukeyHSD(anova3)
```


##T-tests for binary variables
```{r ttest}
#waterfront
waterNo <- saratoga %>%
        select(price, waterfront) %>%
        filter(waterfront == "No")
waterYes <- saratoga %>%
        select(price, waterfront) %>%
        filter(waterfront == "Yes")
t.test(waterNo$price, waterYes$price, alternative = "two.sided")

#newConstruction
constNo <- saratoga %>%
        select(price, newConstruction) %>%
        filter(newConstruction == "No")
constYes <- saratoga %>%
        select(price, newConstruction) %>%
        filter(newConstruction == "Yes")
t.test(constNo$price, constYes$price, alternative = "two.sided")

#centralAir
centNo <- saratoga %>%
        select(price, centralAir) %>%
        filter(centralAir == "No")
centYes <- saratoga %>%
        select(price, centralAir) %>%
        filter(centralAir == "Yes")
t.test(centNo$price, centYes$price, alternative = "two.sided")
```

###Model fitting
#Split the set into train and test
```{r Prepare set}
set.seed(2017)
train.size <- 0.8
train.index <- sample.int(length(saratoga$price), round(length(saratoga$price) * train.size))
train <- saratoga[train.index,]
test <- saratoga[-train.index,]
```


##OLS Linear Regression
```{r Model Fitting}
#log transform
#train$price <- log(train$price + 1)
#train$lotSize <- log(train$lotSize + 1)
#train$livingArea <- log(train$livingArea + 1)
#train$rooms <- log(train$rooms + 1)
#train$age <- log(train$age + 1)

#Linear regression model
##R^2 = 0.6607 before log
##R^2 = 0.5927 after log
model <- lm(price~lotSize+age+landValue+livingArea+pctCollege+
                        bedrooms+fireplaces+bathrooms+rooms+heating+fuel+
                        sewer+waterfront+newConstruction+centralAir, data = train)

summary(model)


#Residual plots
plot(model)
```


##Prediction
```{r pred}
train$pred.price <- predict(model, newdata = train, select =
                                    c(price,lotSize,age,landValue,livingArea,
                                        bedrooms,fireplaces+bathrooms,rooms,heating+fuel+
                                        sewer+waterfront,newConstruction,centralAir))
test$pred.price <- predict(model, newdata = test, select =
                                    c(price,lotSize,age,landValue,livingArea,
                                      bedrooms,fireplaces+bathrooms,rooms,heating+fuel+
                                      sewer+waterfront,newConstruction,centralAir))

#train set 
train.corr <- round(cor(train$pred.price, train$price), 2)
train.RMSE <- round(sqrt(mean((train$pred.price - train$price)^2)))
train.MAE <- round(mean(abs(train$pred.price - train$price)))
c(train.corr^2, train.RMSE, train.MAE)

#test set
test.corr <- round(cor(test$pred.price, test$price), 2)
test.RMSE <- round(sqrt(mean((test$pred.price - test$price)^2)))
test.MAE <- round(mean(abs(test$pred.price - test$price)))
c(test.corr^2, test.RMSE, test.MAE)
```


##Cutoff of extreme values using Cooks distance
```{r cook}
cutoff <- 4/((nrow(train)-length(model$coefficients)-2)) #D plot 4/(n-k-1)
plot(model, which = 4, cook.levels = cutoff) #D values > cutoff
plot(model, which = 5, cook.levels = cutoff)

train <- train[-which(rownames(train)
                      %in% c("602", "1202", "1279")),]
```


##Refit the model
```{r refit}
model <- lm(price~lotSize+age+landValue+livingArea+pctCollege+
                    bedrooms+fireplaces+bathrooms+rooms+heating+fuel+
                    sewer+waterfront+newConstruction+centralAir, data = train)   

summary(model)
```